---
title: "1_BuildMaster.Rmd"
author: "Samuel Gurr"
date: "2026-03-12"
output: html_document
---


## Build master file

**Objective**: Merge three datasets that cover the Yaquina dataset from the 1980s to 2025
data are in different formats as .tsc, zip folder, and .xlsx

#### load required pacakges

* note: use install.packages("") if you do not have these downloaded!

```{r, include=FALSE}
library(lubridate)
library(dplyr)
library(readxl)
library(data.table)
```

#### set your working directory 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# SET WORKING DIRECTORY 
knitr::opts_knit$set(root.dir = "C:/Users/gurrs/Documents/6_Github_repositories/GurrLab/SeaWater-Yaquina/raw") # Sam's work computer
```


### Import data 

  * data being the outputs from the acidifcation, hypoxia, low salinity, high temperature events R code

```{r import data}

# Yaquina data from 2007-2013
Yaquina_2007_2013 <- fread("LOBO0019-20251208164806.tsv", header = TRUE)

# Yaquina data from 1988 - 2004
file_connection   <- unz("OSU_Seawater_88to04.zip", "OSU_Seawater_88to04.csv")
Yaquina_1998_2004 <- read.csv(file_connection)

# Yaquina Sonde from 2009 - 2025
Yaquina_2009_2025 <- read_excel("Mochon Collura-YSI Yaquina 2009_2025-dataset final.xlsx", sheet = 3)

```


## Edit dataframes in prep to merge

* before we merge the three datasets, we first need to do the following: 
  
  * run some QC, truncate outliers and output them
  
  * rename columns, assign NAs if needed 


#### Edit Yaquina_1998_2004
```{r Edit Yaquina_1998_2004, include=FALSE}

# Pad with leading zeros to a total width of 6
# why? the format for the date does not incude '00' for year 2000 nor '0' to start 2001-2004
Yaquina_1998_2004$Date_yymmdd <- sprintf("%06d",Yaquina_1998_2004$Date_yymmdd)
Yaquina_1998_2004$Time_hhmm   <- sprintf("%04d",Yaquina_1998_2004$Time_hhmm)

# format the date as yyyymmdd
Yaquina_1998_2004$Date_yyyymmdd      <- format(as.Date(as.character(Yaquina_1998_2004$Date_yymmdd), format = "%y%m%d"), "%Y%m%d")

# format the date as yyyymmdd_hhmm
Yaquina_1998_2004$Date_yyyymmdd_hhmm <- paste0(Yaquina_1998_2004$Date_yyyymmdd, "_", Yaquina_1998_2004$Time_hhmm)


Yaquina_1998_2004_edit <- as.data.frame(Yaquina_1998_2004) %>% 
                            # reformat date to GMT, and numeric
                            mutate(Date_GMT = lubridate::ymd_hm(Yaquina_1998_2004$Date_yyyymmdd_hhmm, tz = "GMT"),
                                   Date_NUM_FORMAT = (as.numeric(Date_GMT) / 86400 / 365) ) %>% # get a numeric version of time
                            # convert units to universal format
                            dplyr::mutate(Temperature_Kelvin = WaterTemp_YBay_C + 273.15, # from Celcius to Kelvin
                                          Dissolved_oxygen_molm3 = NA, # from mg/L to mol per m⁻³
                                          Dissolved_oxygen_mgL = NA,
                                          pH = NA,
                                          Total_chlorophyll_kgm3 = NA,
                                          Total_chlorophyll_ugL = NA,
                                          Turbidity = NA,
                                          Depth_m = NA) %>% # from ug/L to kg m⁻³
                            # rename columns 
                            dplyr::rename(Temperature_Celcius = WaterTemp_YBay_C,
                                          Salinity = Salinity_YBay_ppt) %>% 
                            # select target columns
                            dplyr::select(Date_GMT, 
                                          Date_NUM_FORMAT, 
                                          Temperature_Kelvin,
                                          Temperature_Celcius,
                                          pH,
                                          Salinity,
                                          Dissolved_oxygen_molm3,
                                          Dissolved_oxygen_mgL,
                                          Total_chlorophyll_kgm3,
                                          Total_chlorophyll_ugL,
                                          Turbidity,
                                          Depth_m) %>% 
                            # add column as the category for the dataset
                            dplyr::mutate(dataset = 'Dataset1_1998_2004',
                                          Location = 'Yaquina_Bay',
                                          Latitude = NA,
                                          Longitute = NA)
```

### Edit Yaquina_2007_2013
```{r Edit Yaquina_2007_2013, include=FALSE}

Yaquina_2007_2013_edit <- as.data.frame(Yaquina_2007_2013) %>% 
                            # reformat date to GMT, and numeric
                            mutate(Date_GMT = lubridate::as_datetime(`date [PST]`),
                                   Date_NUM_FORMAT = (as.numeric(Date_GMT) / 86400 / 365) ) %>% # get a numeric version of time
                            # convert units to universal format
                            dplyr::mutate(Temperature_Kelvin = `temperature [C]` + 273.15, # from Celcius to Kelvin
                                          Dissolved_oxygen_molm3 = `dissolved oxygen [ml/l]` * 0.03125, # from mg/L to mol per m⁻³
                                          pH = NA,
                                          Total_chlorophyll_kgm3 = NA,
                                          Total_chlorophyll_ugL = NA,
                                          Turbidity = NA,
                                          Depth_m = NA) %>% # from ug/L to kg m⁻³
                            # rename columns 
                            dplyr::rename(Temperature_Celcius = `temperature [C]`,
                                          Salinity = salinity,
                                          Dissolved_oxygen_mgL = `dissolved oxygen [ml/l]`) %>% 
                            # select target columns
                            dplyr::select(Date_GMT, 
                                          Date_NUM_FORMAT, 
                                          Temperature_Kelvin,
                                          Temperature_Celcius,
                                          pH,
                                          Salinity,
                                          Dissolved_oxygen_molm3,
                                          Dissolved_oxygen_mgL,
                                          Total_chlorophyll_kgm3,
                                          Total_chlorophyll_ugL,
                                          Turbidity,
                                          Depth_m) %>% 
                            # add column as the category for the dataset
                            dplyr::mutate(dataset = 'Dataset2_2007_2013',
                                          Location = 'Yaquina_Bay',
                                          Latitude = NA,
                                          Longitute = NA)
```

#### Edit Yaquina_2009_2025
```{r Edit Yaquina_2009_2025, include=FALSE}

# flag label in the file suggests that "1" == reliabe,
# isolating the values 2,3 and outputing them if future users want to view and QC
Yaquina_2009_2025_flagged <- as.data.frame(Yaquina_2009_2025) %>% dplyr::filter(Flags > 1)


Yaquina_2009_2025_edit <- as.data.frame(Yaquina_2009_2025) %>% 
                            # flags == 1, Pass
                            dplyr::filter(Flags == 1) %>%  
                            # reformat date to GMT, and numeric
                            mutate(Date_GMT = lubridate::as_datetime(Timestamp.GMT),
                                   Date_NUM_FORMAT = (as.numeric(Date_GMT) / 86400 / 365) ) %>% # get a numeric version of time
                            # convert units to universal format
                            dplyr::mutate(Temperature_Kelvin = Temperature.C + 273.15, # from Celcius to Kelvin
                                          Dissolved_oxygen_molm3 = ODO.mgL * 0.03125, # from mg/L to mol per m⁻³
                                          Total_chlorophyll_kgm3 = Chlorophyll /(1000000)) %>% # from ug/L to kg m⁻³
                            # rename columns 
                            dplyr::rename(Temperature_Celcius = Temperature.C,
                                          Depth_m = Depth.m,
                                          Dissolved_oxygen_mgL = ODO.mgL,
                                          Total_chlorophyll_ugL = Chlorophyll) %>% 
                            # select target columns
                            dplyr::select(Date_GMT, 
                                          Date_NUM_FORMAT, 
                                          Temperature_Kelvin,
                                          Temperature_Celcius,
                                          pH,
                                          Salinity,
                                          Dissolved_oxygen_molm3,
                                          Dissolved_oxygen_mgL,
                                          Total_chlorophyll_kgm3,
                                          Total_chlorophyll_ugL,
                                          Turbidity,
                                          Depth_m) %>% 
                            # add column as the category for the dataset
                            dplyr::mutate(dataset = 'Dataset3_2009_2025',
                                          Location = 'Yaquina_Bay',
                                          Latitude = NA,
                                          Longitute = NA)


```



## Merge 'Master' file

```{r, include=FALSE}
Master <- as.data.frame(
                      rbind(Yaquina_1998_2004_edit,
                            Yaquina_2007_2013_edit,
                            Yaquina_2009_2025_edit)
                      )

View(Master)
```

## Output 'Master' file

```{r, include=FALSE}

readr::write_csv(Master, "../RAnalysis/Output/Compiled_master_file.csv.gz")
```

